Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-03-07 draft.tex
.

\subsubsection{The Role of Firing Rate Heterogeneity---a Dichotomy between Heterogeneity and Control?}
Modeling homeostasis necessarily includes a form of error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. Of course this raises the question of how to choose these homeostatic targets. Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. Many statistical analyses found particular evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity.

@@

Interestingly, measurements in the barrel cortex and hippocampus have found that firing rates during exploration or task performance are significantly correlated with those found for spontaneous, baseline activity \cite{OConnor_2010,Mizuseki_2013}. This suggests that the same set of neurons tend to be relatively active or inactive under different stimuli and changing environments, in line with the idea of modeling homeostasis by means of individual targets of activity, which together form the desired statistical distribution. Still, though apparently being a good phenomenological approach, it is unclear what gives rise to these target differences.

Another problem that one might consider is the fact that temporal fluctuations of activity are believed to be a major part of information transmission and processing within the brain \cite[p. 123--150]{Theor_Neur_Dayan}. Homeostasis as described above could potentially override information by eliminating these fluctuations of activity. However, it is argued that homeostatic mechanisms act on time scales of hours to days and thus do not interfere with changes in activity taking place on the order of seconds or below \cite{Turrigiano_2011}. Returning to the previous question, though the slowness of homeostatic adaptation might allow for temporal fluctuations of activity, it does not overcome the conceptual dichotomy between single cell homeostasis and persistent variability in spontaneous activity. Marder and Goaillard speculated that instead of a mechanism controlling activity on a single-cell level, one might consider the possibility of a ``global sensor of network performance to provide error signals for tuning of the network" \cite{Marder_2006}. This concept was used by Sweeney et al. as the basic idea for the introduction of a diffusive, intercellular homeostatic mechanism, which will be introduced in the next section.

@@

\subsubsection{Synaptic Lifetimes}
Synaptic lifetimes have been shown to approximately follow a power law distribution in earlier versions of LIF-SORN and binary SORN \cite{SORN_Paper,Pengsheng_2013}. This feature was preserved under diffusive homeostasis, as shown in Figure \ref{Syn_Lifetimes}. The resulting slopes are within previously reported values of $\mathrm{\approx - 5/3}$ \cite{SORN_Paper}.
\begin{figure}
%%% plot_syn_lifetimes.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_lifetimes.png}
\caption{Distribution of lifetimes of recurrent excitatory synaptic connections. Linear fits ignoring falloff (only using bins with $\rm t \leq 150 \, s$) for large synaptic lifetimes resulted in a slope of $\mathrm{-1.87}$ for diffusive homeostasis and $\mathrm{-1.68}$ for non-diffusive homeostasis.}
\label{Syn_Lifetimes}
\end{figure}
To get an approximative explanation for the observed slope in the region of small lifetimes, we note that short synaptic lifetimes are most likely coming from synapses whose weights did not reach large values compared to the initial weight. The effect of multiplicative normalization is small compared to additive STDP in this regime of small weights. Small weights are therefore determined by additive quasi-random (due to their weak coupling to the postsynaptic neuron) fluctuations, which, as a very simple approximation, can be described as a Brownian motion with a certain standard deviation $\mathrm{\sigma}$. The first-hitting time of this Brownian motion is described by a Lévy distribution \cite{Roy_1968}:
\begin{equation}
p(t,w_0,w_{\rm prune}) = \frac{ | w_0-w_{\rm prune}|}{\sqrt{2\pi \sigma t^3}} \mathrm{exp} \left( \frac{ \left( w_0-w_{\rm prune} \right) ^2}{2\sigma t}\right)
\label{Levy_Dist}
\end{equation}
The term $\mathrm{t^{-3/2}}$ dominates for larger $\mathrm{t}$, which very roughly resembles the power-law exponents found in Figure \ref{Syn_Lifetimes}. Zheng et al. found a similar slope in the binary version of the SORN and also argued that this could be referred to random-walk behavior \cite{Pengsheng_2013}. Still, understanding the exponent of this distribution is subject to ongoing research and it has been argued that a more precise explanation could be used to relate certain parameters of synaptic plasticity to the value of this exponent. Especially, the balance between long-term potentiation and depression is thought to influence this exponent, where potentiation/depression dominated plasticity leads to a smaller/larger absolute value of the exponent of the distribution.

>>>>

\subsubsection{Former Single-Cell Homeostasis as an Integral Control}
Former versions of the LIF-SORN and SORN used a simple form of integral control to regulate excitatory firing rates. Individual firing rates of excitatory cells were estimated by a spike counter over short subsequent time periods and compared to a target value. This difference was then fed to a controller which increased or decreased neuronal excitability in an additive way (making it a discrete form of integral control). As mentioned before, this type of control allowed to reliably tune each cell to a given target firing rate under changing input. A global target firing rate was chosen for all excitatory neurons. Consequently, this resulted in a very narrow distribution of excitatory firing rates.

\subsubsection{The Role of Firing Rate Heterogeneity---a Dichotomy between Heterogeneity and Control?}
Modeling homeostasis necessarily includes a form of error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. Of course this raises the question of how to choose these homeostatic targets. As mentioned in the previous section, former versions of the SORN defined a single global target value. However, experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. Many statistical analyses found particular evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, claim that it is reasonable to define a single global target of activity.

@@

Interestingly, measurements in the barrel cortex and hippocampus have found that firing rates during exploration or task performance are significantly correlated with those found for spontaneous, baseline activity \cite{OConnor_2010,Mizuseki_2013}. This suggests that the same set of neurons tend to be relatively active or inactive under different stimuli and changing environments, in line with the idea of modeling homeostasis by means of individual targets of activity, which together form the desired statistical distribution. Similar findings have recently been reported by Hengen et al. \cite{Hengen_2016}. They measured firing rates of individual neurons in the primary visual cortex of rats over several days after monocular deprivation. Among others, one of the results was that after an initial decline of firing rates due to the continuous lack of visual stimuli, firing rates subsequently returned to pre-deprivation levels on a single-cell level within an accuracy of 15\%, on average. However, they also reported that this accommodation took place predominantly during active awake phases, while firing rates remained relatively unchanged during sleep. This, however, does not attenuate that their result generally supports the existence of single-cell firing rate targets.

Still, though apparently being a good phenomenological description, it is unclear what gives rise to these target differences. Another problem that one might consider is the fact that temporal fluctuations of activity are believed to be a major part of information transmission and processing within the brain \cite[p. 123--150]{Theor_Neur_Dayan}. Homeostasis as described above could potentially override information by eliminating these fluctuations of activity. However, it is argued that homeostatic mechanisms act on time scales of hours to days and thus do not interfere with changes in activity taking place on the order of seconds or below \cite{Turrigiano_2011}. Returning to the previous question, though the slowness of homeostatic adaptation might allow for temporal fluctuations of activity, it does not overcome the conceptual dichotomy between single cell homeostasis and persistent variability in spontaneous activity. Marder and Goaillard speculated that instead of a mechanism controlling activity on a single-cell level, one might consider the possibility of a ``global sensor of network performance to provide error signals for tuning of the network" \cite{Marder_2006}. This concept was used by Sweeney et al. as the basic idea for the introduction of a diffusive, intercellular homeostatic mechanism, which will be introduced in the next section.

@@

\subsubsection{Synaptic Lifetimes}
Synaptic lifetimes have been shown to approximately follow a power law distribution in earlier versions of LIF-SORN and binary SORN \cite{SORN_Paper,Pengsheng_2013}. Power law distributions appear in different areas of natural and social sciences \cite{Gabaix_1999,Pascual_2002,Kagan_2010}. Some quantity $x$ is said to be power law distributed if its distribution $p(x)$ obeys
\begin{equation}
p(x) \propto x^{-\alpha}
\label{Pow_Law}
\end{equation}
with an exponent $\alpha>0$. Note that for this distribution to be properly normalized, one needs to set a lower (upper) bound $x_{\rm min.}$($x_{\rm max.}$) if $\alpha \geq 1$ ($\alpha \leq 1$). This family of distributions is of particular interest with respect to the fact that the ``long tail" of the distribution contains a significant portion of the total normalized integral of the probability density function. In other words, the probability of the appearance of exceptionally large values of $x$ can not be regarded as insignificant. Analyses of empirical data that is believed to obey a power-law distribution thus usually focus on the statistics of rare large events.

In our case of synaptic lifetimes, this raises a practical problem: Measuring lifetimes over a finite total simulation time leads to a fallow of probability for larger lifetimes. Furthermore, synapses whose lifetimes \emph{could} have been potentially longer but were simply terminated by the end of the simulation will cause a bunching and over-representation of lifetimes within the time range given by the total simulation time. There exists no general solution for the first problem other than running longer simulations. The second problem could be tackled by only including synapses into the statistics which terminated before the end of the simulation, though this of course reduces the size of the dataset. Practically, we accounted for this by only including lifetime periods into the statistics that explicitly contained an ``on-off" transition in the end.

Another question regarding the data

Still, estimating an power-law exponent based on the given data required setting an upper and lower limit over which

 This feature was preserved under diffusive homeostasis, as shown in Figure \ref{Syn_Lifetimes}. The resulting slopes are within previously reported values of $\mathrm{\approx - 5/3}$ \cite{SORN_Paper}.
\begin{figure}
%%% plot_syn_lifetimes.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_lifetimes.png}
\caption{Distribution of lifetimes of recurrent excitatory synaptic connections. Linear fits ignoring falloff (only using bins with $\rm t \leq 150 \, s$) for large synaptic lifetimes resulted in a slope of $\mathrm{-1.87}$ for diffusive homeostasis and $\mathrm{-1.68}$ for non-diffusive homeostasis.}
\label{Syn_Lifetimes}
\end{figure}
To get an approximative explanation for the observed slope in the region of small lifetimes, we note that short synaptic lifetimes are most likely coming from synapses whose weights did not reach large values compared to the initial weight. The effect of multiplicative normalization is small compared to additive STDP in this regime of small weights. Small weights are therefore determined by additive quasi-random (due to their weak coupling to the postsynaptic neuron) fluctuations, which, as a very simple approximation, can be described as a Brownian motion with a certain standard deviation $\mathrm{\sigma}$. The first-hitting time of this Brownian motion is described by a Lévy distribution \cite{Roy_1968}:
\begin{equation}
p(t,w_0,w_{\rm prune}) = \frac{ | w_0-w_{\rm prune}|}{\sqrt{2\pi \sigma t^3}} \mathrm{exp} \left( \frac{ \left( w_0-w_{\rm prune} \right) ^2}{2\sigma t}\right)
\label{Levy_Dist}
\end{equation}
The term $\mathrm{t^{-3/2}}$ dominates for larger $\mathrm{t}$, which very roughly resembles the power-law exponents found in Figure \ref{Syn_Lifetimes}. Zheng et al. found a similar slope in the binary version of the SORN and also argued that this could be referred to random-walk behavior \cite{Pengsheng_2013}. Still, understanding the exponent of this distribution is subject to ongoing research and it has been argued that a more precise explanation could be used to relate certain parameters of synaptic plasticity to the value of this exponent. Especially, the balance between long-term potentiation and depression is thought to influence this exponent, where potentiation/depression dominated plasticity leads to a smaller/larger absolute value of the exponent of the distribution.
